---
title: "CEAP Analysis"
author: "José Benardi de Souza Nunes"
output:
  html_notebook:
    toc: yes
    toc_float: yes
---

```{r}
library(tidyverse)
library(here)
```

```{r}
data <- read_csv(here::here("data/dadosCEAP.csv"),
                 progress = F,
                 col_types = cols(
                   nomeParlamentar = col_character(),
                   idCadastro = col_integer(),
                   sgUF = col_character(),
                   sgPartido = col_character(),
                   tipoDespesa = col_character(),
                   especDespesa = col_character(),
                   fornecedor = col_character(),
                   CNPJCPF = col_character(),
                   tipoDocumento = col_integer(),
                   dataEmissao = col_character(),
                   valorDocumento = col_double(),
                   valorGlosa = col_integer(),
                   valorLíquido = col_double())) %>%
  mutate(dataEmissao = lubridate::parse_date_time(dataEmissao,
                                                  "%Y-%m-%d %H:%M:%S"),
         year_month = paste(lubridate::year(dataEmissao),
                            lubridate::month(dataEmissao),
                            sep = "-"),
         tipoDespesa = toupper(tipoDespesa))

state_info <- read_csv(here::here("/data/limiteMensalCEAP.csv"),
                 progress = F,
                 col_types = cols(
                   UF = col_character(),
                   limite_mensal = col_double()))

data %>% 
  full_join(state_info,
            by = c("sgUF" = "UF")) -> data

data %>%
  glimpse()
```

# Which parliamentarians spent the most from the CEAP budget? Which spent less? 

```{r}
data %>%
  group_by(sgUF,
           idCadastro,
           nomeParlamentar) %>%
  summarize(total_expense = sum(valorLíquido)) %>%
  ungroup() %>%
  top_n(n=10, wt=total_expense) %>%
  ggplot(aes(total_expense, 
             reorder(nomeParlamentar,
                     total_expense))) +
  geom_segment(aes(x = min(total_expense),
                   y =reorder(nomeParlamentar,
                              total_expense),
                   xend = total_expense,
                   yend = nomeParlamentar),
               color = "grey50",
               size=0.35) +
  geom_point(aes(color=sgUF),size=3.5) +
  scale_x_continuous(breaks = seq(from = 135E3,
                                  to=1E7,
                                  by = 1E4)) +
  theme(axis.text.x = element_text(angle = 30,
                                   hjust = 1)) +
  labs(x="Total Expense (R$)", 
       y="Parliamentarian name")
```

```{r}
data %>%
  group_by(sgUF,
           idCadastro,
           nomeParlamentar) %>%
  summarize(total_expense = sum(valorLíquido)) %>%
  ungroup() %>%
  filter(total_expense >= 0) %>%
  arrange(total_expense) %>%
  slice(1:10) %>%
  ggplot(aes(total_expense, 
             reorder(nomeParlamentar,total_expense))) +
  geom_segment(aes(x = 0, y =reorder(nomeParlamentar,-total_expense),
                   xend = total_expense,
                   yend = nomeParlamentar),
               color = "grey50",
               size=0.35) +
  geom_point(aes(color=sgUF),size=3.5) +
  scale_x_continuous(breaks = seq(from = 0,
                                  to=1E5,
                                  by = 20)) +
  labs(x="Total Expense (R$)", 
       y="Parliamentarian name")
```

```{r}
data %>%
  filter(tipoDocumento == 2) %>%
  group_by(sgUF) %>%
  summarise(international_expense = sum(valorLíquido)) %>%
  arrange(international_expense)
```

# In which states the parliamentarians spent the most abroad? In which states they spent less? 

```{r}
data %>%
  filter(tipoDocumento == 2) %>%
  group_by(sgUF) %>%
  summarize(international_expense = sum(valorLíquido)) %>%
  ggplot(aes(reorder(sgUF,international_expense),
             international_expense)) +
  geom_bar(stat = "identity") +
  labs(x="States", y="International expense")
```

# Which parties contain the parliamentarians that used CEAP more frequently in the state of Paraíba? Which contain the ones which used less frequently? And in terms of R$?

```{r}
data %>%
  filter(sgUF == "PB") %>%
  na.omit(sgUF,
          sgPartido) %>%
  group_by(sgPartido) %>%
  summarize(frequency =n()) %>%
  ggplot(aes(reorder(sgPartido,frequency),
             frequency)) +
  geom_bar(stat = "identity") +
  labs(x="Parties", y="Absolute frequency")
```

```{r}
data %>%
  filter(sgUF == "PB") %>%
  na.omit(sgUF,
          sgPartido,
          valorLíquido) %>%
  group_by(sgPartido) %>%
  summarize(total_expense = sum(valorLíquido)) %>%
  ggplot(aes(reorder(sgPartido,total_expense),
             total_expense)) +
  geom_bar(stat = "identity") +
  labs(x="Parties", y="Total expense")
```

# Which parliamentars violate the CEAP budget set by their state the most?

```{r}
data %>%
  group_by(sgUF,
           idCadastro,
           limite_mensal,
           nomeParlamentar,
           year_month) %>%
  summarise(monthly_expense = sum(valorLíquido)) %>%
  filter(year_month != "NA-NA") %>%
  filter(monthly_expense > limite_mensal) %>%
  ungroup() %>%
  group_by(idCadastro,
           nomeParlamentar) %>%
  arrange(desc(monthly_expense)) %>% 
  slice(1) %>%
  ungroup() %>%
  top_n(n=10, wt=monthly_expense) %>%
  mutate(overbudget = monthly_expense - limite_mensal) %>%
  ggplot(aes(overbudget,
             reorder(nomeParlamentar,
                     overbudget))) +
  geom_segment(aes(x = 1E5, 
                   y =reorder(nomeParlamentar,
                              overbudget),
                   xend = overbudget,
                   yend = nomeParlamentar),
               color = "grey50",
               size=0.35) +
  geom_point(aes(color=sgUF),size=3) +
  labs(x="Overbudget",
       y="Parliamentar")

```

# In which states the parliamentarians spent the most abroad? In which states they spent less? 

```{r}
data %>%
  filter(tipoDespesa == "PASSAGENS AÉREAS") %>%
  group_by(sgUF) %>%
  summarize(international_expense = sum(valorLíquido)) %>%
  filter(sgUF != "NA") %>%
  ggplot(aes(reorder(sgUF,international_expense),
             international_expense)) +
  geom_bar(stat = "identity") +
  labs(x="States", y="Flight expense")
```

# What type of expense is more frequently used with CEAP by parliamentarians of these parties? What type of expense is more costly?

## PSDB

```{r}
data %>%
  filter(sgPartido %in% c("PSDB")) %>%
  group_by(tipoDespesa) %>%
  summarize(frequency = n()) %>%
  ggplot(aes(reorder(tipoDespesa,frequency),
             frequency)) +
  geom_bar(stat="identity",
           fill="#1E347F") + 
  coord_flip()
```

```{r}
data %>%
  filter(sgPartido %in% c("PSDB")) %>%
  group_by(tipoDespesa) %>%
  summarize(expense = sum(valorLíquido)) %>%
  ggplot(aes(reorder(tipoDespesa,expense),
             expense)) +
  geom_bar(stat="identity",
           fill="#1E347F") +
  labs(y= "Total expense", 
       x="Type of expense") +
  coord_flip() 
```

## PMDB

```{r}
data %>%
  filter(sgPartido %in% c("PMDB")) %>%
  group_by(tipoDespesa) %>%
  summarize(frequency = n()) %>%
  ggplot(aes(reorder(tipoDespesa,frequency),
             frequency)) +
  geom_bar(stat="identity",
           fill="#7F6607") + 
  labs(y= "Absolute Frequency", 
       x="Type of expense") +
  coord_flip()
```

```{r}
data %>%
  filter(sgPartido %in% c("PMDB")) %>%
  group_by(tipoDespesa) %>%
  summarize(expense = sum(valorLíquido)) %>%
  ggplot(aes(reorder(tipoDespesa,expense),
             expense)) +
  geom_bar(stat="identity",
           fill="#7F6607") +
  labs(y= "Total expense", 
       x="Type of expense") +
  coord_flip() 
```

## PT

```{r}
data %>%
  filter(sgPartido %in% c("PT")) %>%
  group_by(tipoDespesa) %>%
  summarize(frequency = n()) %>%
  ggplot(aes(reorder(tipoDespesa,frequency),
             frequency)) +
  geom_bar(stat="identity",
           fill="#7F1707") + 
  labs(y= "Absolute Frequency", 
       x="Type of expense") +
  coord_flip()
```

```{r}
data %>%
  filter(sgPartido %in% c("PT")) %>%
  group_by(tipoDespesa) %>%
  summarize(expense = sum(valorLíquido)) %>%
  ggplot(aes(reorder(tipoDespesa,expense),
             expense)) +
  geom_bar(stat="identity",
           fill="#7F1707") +
  labs(y= "Total expense", 
       x="Type of expense") +
  coord_flip() 
```
