---
title: "C.E.A.P analysis (suppliers and weekend expenses)"
subtitle: "Exploratory data analysis on 'Quota for the exercise of parliamentary activity' provided by the Brazilian government"
author: "José Benardi de Souza Nunes"
date: "06/09/2018"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
  html_notebook:
    toc: yes
    toc_float: yes
---

<br>

# Introduction 

<br>

> This exploratory data analysis was made based on data provided by the Brazilian government  about the expenses allowed to its parliamentarians or C.E.A.P. (Cota para o Exercício da Atividade Parlamentar / Quota for the Exercise of Parliamentary Activity). More information about it (in Portuguese) can be found in its [official site](http://www2.camara.leg.br/transparencia/acesso-a-informacao/copy_of_perguntas-frequentes/cota-para-o-exercicio-da-atividade-parlamentar)

<br>

***

<br>

# Data Overview 

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
library(scales)
library(lattice)
library(ggridges)
library(tidyverse)
library(lubridate)

theme_set(theme_bw())
```

```{r}
data <- read_csv(here::here("data/dadosCEAP.csv"),
                 progress = F,
                 col_types = cols(
                   nomeParlamentar = col_character(),
                   idCadastro = col_integer(),
                   sgUF = col_character(),
                   sgPartido = col_character(),
                   tipoDespesa = col_character(),
                   especDespesa = col_character(),
                   fornecedor = col_character(),
                   CNPJCPF = col_character(),
                   tipoDocumento = col_integer(),
                   dataEmissao = col_character(),
                   valorDocumento = col_double(),
                   valorGlosa = col_integer(),
                   valorLíquido = col_double())) 

data %>% 
  mutate(dataEmissao = parse_date_time(dataEmissao,"%Y-%m-%d %H:%M:%S"),
         year_month = paste(lubridate::year(dataEmissao),   # extract year
                            lubridate::month(dataEmissao),  # extract month
                            sep = "-"),
         tipoDespesa = toupper(tipoDespesa)) -> data

state_info <- read_csv(here::here("/data/limiteMensalCEAP.csv"),
                       progress = F,
                       col_types = cols(
                         UF = col_character(),
                         limite_mensal = col_double()))

data %>% 
  full_join(state_info,
            by = c("sgUF" = "UF")) -> data

data %>%
  glimpse()
```

<br>

```{r}
read_tsv(here::here("data/trsl_expenses.tsv"),
         progress = F,
         col_types = cols(
           original_term = col_character(),
           translation = col_character())) -> trsltd_expenses

trsltd_expenses
```

* A translation for the description of each one of the 18 types of expense was recorded alongside its original term.

<br>

```{r}
data %>% 
  full_join(trsltd_expenses,
            by = c("tipoDespesa" = "original_term"))  %>% 
  mutate(tipoDespesa = translation) %>%  # translate col tipoDespesa 
  select(-translation)-> data # drop unused column

data %>%
  sample_n(5) %>%
  select(nomeParlamentar,
         tipoDespesa,
         dataEmissao)
```

* We now have each term in the column tipoDespesa translated.

<br>

```{r}
data %>%
  mutate(week_day = wday(dataEmissao), # extract week day
         is_weekend = (week_day == 1) |  # deduce whether it's a weekend
                         (week_day == 7)) -> data

data %>%
  select(dataEmissao,
         week_day,
         is_weekend) %>%
  sample_n(5)
```

* Once we extracted which day of the week the emission was made we simply deduced whether it was a weekend or not.

<br>

```{r}
data %>%
  na.omit(year_month) %>% 
  group_by(year_month) %>%     # times used
  summarize(total = n()) %>%   # by year/month
  ggplot(aes(year_month, total)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 60,
                                   hjust = 1)) +
  labs(x="Year and month", y="Absolute Frequency")
```

* The period of time covered by the dataset is of about 3 years which amounts to the duration of our parliamentarian's office minus the 120 days prior to the election.

<br>

```{r}
data %>%
  group_by(idCadastro,
           nomeParlamentar) %>%
  slice(1) %>% # keep one entry for parliamentarian
  ungroup() %>%
  group_by(sgPartido) %>%              # calc parliamentarian  
  summarize(num_parliament = n()) %>%  # by party
  na.omit() %>%
  ggplot(aes(reorder(sgPartido,
                     num_parliament),
             num_parliament)) +
  geom_bar(stat="identity") + 
  labs(x="Political parties",
       y="Number of parliamentarians") +
  theme(axis.text.x = element_text(angle = 60,
                                   hjust = 1))
```

* We can see that **PMDB** detains the majority of offices followed closely by **PT** and that **PSDB** is still a major player. This was to be expected as the aforementioned parties are often regarded as the most prominent ones in the Brazilian political scene.  

<br>

***

<br>

# Expenditure during the week and the weekends

<br>

```{r}
data %>%
   mutate(month = lubridate::month(dataEmissao)) %>%  # extract month
  na.omit(week_day,month) %>% 
  group_by(week_day,month) %>% # times used
  summarize(total = sum(valorLíquido)) %>%  # by day of week / month 
  tidyr::spread(key= week_day, value=total, 
                fill = 0) %>%
  select(-month) %>% data.matrix() -> matrix_yearly

colnames(matrix_yearly) <- c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")

x.scale <- list(at=seq(from=1,to=12,by=1))

levelplot(matrix_yearly, 
          col.regions=heat.colors,
          scales=list(x=x.scale),
          main="Total expenditure across days of the week",
          xlab="month of the year", ylab= "week day")
```

* Overall we can see a clear predominance of expenditures during the week as both **saturday** and **sunday** have a smaller  total expenditure across the whole year.
    + It was to be expected, as CEAP is meant to be used to pay for work related expenses.   

<br>

```{r}
data %>%
  group_by(is_weekend,sgPartido) %>%
  summarize(total_cost = sum(valorLíquido)) %>%
  na.omit(is_weekend) -> party_expense

party_expense %>%
  ungroup() %>%
  sample_n(5)
```

* Now we have the total cost incurred by each party during the week and during the weekend. 

<br>

```{r}
ggplot(data= party_expense,
       aes(reorder(sgPartido,
                     total_cost),
             total_cost,
             fill=is_weekend)) +
  geom_bar(data=subset(party_expense,
                       !is_weekend),
           stat = "identity") + 
  geom_bar(data=subset(party_expense,
                       is_weekend),
           aes(y=-total_cost),
           stat = "identity") +
  coord_flip() +
   labs(x="Political parties",
        y="Total expenditure (R$)")
```

* Across different political parties the strong predominance of expenditures during the week holds.
    + PMDB is the biggest spender.
* **PMDB**, **PT** and **PSDB** are the parties with the highest number of offices, therefore it's reasonable that they are the ones to spend the most. 
    
<br>

## Which parties spent the most during the weekend?

```{r}
data %>%
  filter(is_weekend) %>%
  group_by(is_weekend,sgPartido) %>%
  summarize(total_cost = sum(valorLíquido)) %>%
  na.omit(is_weekend) %>%
  ggplot(aes(reorder(sgPartido,
                     total_cost),
             total_cost)) +
  geom_bar(stat="identity",
           fill="#00BFC4") + 
  labs(x="Political parties",
       y="Total expenditure (R$)",
       title="Expenditures during the weekend") +
  theme(axis.text.x = element_text(angle = 60,
                                   hjust = 1))
```

* **PMDB**, **PT** and **PSDB** are also the ones to spend the highest amount of money during the weekend. 
* Despite the fact that overall **PMDB** is the biggest spender (as seen before) **PT** spent the most during the weekend.

> **PMDB**, **PT** and **PSDB**  spent the most during the weekend.

<br>

## Which parties had most of their expenditures made during the weekend? 

```{r}
data %>%
  na.omit(is_weekend,
          sgPartido) %>%
  group_by(sgPartido) %>%
  mutate(total = sum(valorLíquido)) %>% # calc total cost by party
  ungroup() %>%
  group_by(total,
           sgPartido,
           is_weekend) %>%                        # calc cost by weekdays
  summarize(total_cost = sum(valorLíquido)) %>%   # and by weekends
  mutate(fract = total_cost/total) %>%          # calc fraction
  filter(is_weekend) %>% 
  ggplot(aes(reorder(sgPartido,fract),
             fract)) +
  geom_bar(stat="identity",
           fill="#00BFC4") +
    labs(x="Political parties",
       y="Percentage (%)",
       title="Expenditures during the weekend in R$") +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 60,
                                   hjust = 1))
```

* Unexpectedly, we see as leader the party **PRTB** (ironically named Brazilian Labour Renewal Party) that made almost $40\%$ of its expenditures during the weekend.
* **PT** not only spent a great amount of money during the weekend as seen before, but had a substantial fraction of its expenditures made during the weekend.
* We can see that parties like **PV** and **PSC** which were obscured in terms of overall expense, stand out as some of the parties who spent a good part of their budget during the weekend.  

> **PRTB** was the party to have most of its expenditures made during the weekend.

<br>

***

<br>

# Looking at the cost of the suppliers  

<br>

## What were the costliest suppliers (in terms of median)? What services did they offer?  

<br>

#####Let's separate the top 10 most expensive suppliers. They will be selected by their median cost across all parties.

```{r}
data %>%
  mutate(fornecedor = toupper(fornecedor)) %>%
  mutate(fornecedor = ifelse(         # shorten the name       
    fornecedor == "TODDE ADVOGADOS E CONSULTORES ASSOCIADOS", 
    "TODDE ADVOGADOS",
    fornecedor)) %>%
  group_by(fornecedor,
           sgPartido,
           tipoDespesa) %>%              # calc expense of each
  summarize(total_cost = sum(valorLíquido)) %>% # supplier for each party
  ungroup() %>%
  group_by(fornecedor) %>%                           # calc median expense of
  mutate(median_expense = median(total_cost)) %>% # each supplier (all parties)
  ungroup() -> supplier_data

supplier_data %>%
  select(fornecedor,
         median_expense,
         sgPartido,
         total_cost) %>%
  sample_n(5)
```

* Now that we  have the median cost of each supplier we can examine how the top 10 most expensive suppliers relate to the parties.

<br>

```{r}
supplier_data %>%
  select(median_expense) %>%
  unique() %>%
  arrange(desc(median_expense)) %>%
  slice(10) -> tenth_supplier # 10th biggest median cost

supplier_data %>%
  group_by(fornecedor,
           sgPartido,
           tipoDespesa) %>%
  filter(median_expense >= tenth_supplier) %>%  # filter top 10 suppliers
  ggplot(aes(reorder(fornecedor,
                     median_expense),
             total_cost,
             color=tipoDespesa)) +
  geom_boxplot() +
  labs(x="Supplier",
       y="Cost (R$)",
       color=NULL) +
  coord_flip()+
  theme(legend.position="bottom",
        legend.direction = "vertical") +
    theme(legend.text=element_text(size=8),
           axis.text.y = element_text(size=8))
```

<br>

* TAM and GOL which are both air flight companies have the most diverse behavior, accounting for the lowest cost in the top 10 but majoritarily being _the costliest suppliers of all_.
    + In terms of median the air flight companies **TAM** and **GOL** take first and second place as costliest suppliers of all with $\color{red}{\text{Air Ticket Issuance}}$

<br>

* The rest of the suppliers are plainly less erratic.
    + Publicity and advertising companies  such as **STILO GRAFICA** and **GRAFICA SATELITE** take the third and fourth place with $\color{blue}{\text{Divulgation of parliamentary activity}}$.
    + The law firm **Todde advogados e consultores associados** takes the fifth place with $\color{darkgreen}{\text{Consultancies, research and technical works}}$

<br>

> The costliest suppliers of all were **TAM** and **GOL** with _Air Ticket Issuance_



